---
title: "Apple's mobility data"
author:
- name: Kieran Healy
  affiliation: Duke University
  email: kjhealy@soc.duke.edu
date: '`r format(Sys.Date(), "%B %d, %Y")`'
---


Code to accompany [this blog post](https://kieranhealy.org/blog/archives/2020/04/23/apples-covid-mobility-data/). 
If you don't have them, install the required libraries with `install.packages()`. Install the `covdata` package 
by following [the instructions at the package's homepage](https://kjhealy.github.io/covdata).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.showtext = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
library(here)
library(janitor)
library(socviz)
library(ggrepel)
library(covdata)
```

```{r, eval = FALSE}
## --------------------------------------------------------------------
## Custom font and theme. By default this code chunk is not evaluated.
## --------------------------------------------------------------------
library(showtext)
showtext_auto()
library(myriad)
import_myriad_semi()

theme_covid <- function() {
  theme_myriad_semi() +
    theme(plot.title = element_text(size = rel(2), face = "bold"),
          plot.subtitle = element_text(size = rel(1.5)),
          plot.caption = element_text(size = rel(0.9)),
          axis.text.y = element_text(size = rel(2)),
          axis.title.x = element_text(size = rel(1.5)),
          axis.title.y = element_text(size = rel(1.5)),
          axis.text.x = element_text(size = rel(2)),
          legend.text = element_text(size = rel(2))
    )
}


theme_set(theme_myriad_semi())
## --------------------------------------------------------------------
```

```{r}
## --------------------------------------------------------------------
## For the time series decomposition
library(tsibble)
library(feasts)
library(fable)
library(timeDate)
## --------------------------------------------------------------------

## --------------------------------------------------------------------
## Convenience functions
## --------------------------------------------------------------------

## such an ugly function (because it's like 10 years old)
my_colors <- function(palette="cb"){
### The palette with black:
  cb.palette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
  ## Same one Reversed
  rcb.palette <- rev(cb.palette)
  ## Blue and yellow first choices
  bly.palette <- c("#E69F00", "#0072B2", "#000000", "#56B4E9", "#009E73", "#F0E442", "#D55E00", "#CC79A7")
  if (palette=="cb") return(cb.palette) else if (palette=="rcb") return(rcb.palette) else if (palette=="bly") return(bly.palette) else stop("Choose cb, rcb, or bly ony.")
}

is_max <- function(x) {
  seq_along(x) == which.max(x)
}

## Check for figures/ 
ifelse(!dir.exists(here("figures")),
       dir.create(file.path("figures")),
       FALSE)


```


```{r}
head(apple_mobility)
tail(apple_mobility)
```

We're going to rename the `index` column to `score`, because weird things can happen with `dplyr` and especially `tsibble` (and related functions from the `fable` and `feasts` packages) if you have a column named `index`.

```{r}
apple_mobility <- apple_mobility %>%
  rename(score = index)
```


```{r}
## Quick look
p_base_all <- apple_mobility %>%
  filter(geo_type == "city") %>%
  select(region:score) %>%
  rename(mode = transportation_type) %>%
  ggplot(aes(x = date, y = score, group = mode, color = mode)) +
  geom_line(size = 0.5) +
  scale_color_manual(values = my_colors("bly")) +
  facet_wrap(~ region, ncol = 8) +
  labs(x = "Date", y = "Trend",
       color = "Mode",
       title = "All Modes, All Cities, Base Data",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")

ggsave(here("figures/all_cities_raw.png"), p_base_all, width = 20, height = 30, dpi = 300)
```

Create a `tsibble` object of the cities of interest. Apple has expanded the number of cities covered, so we run into the following problem:

```{r}

apple_mobility %>%
  filter(geo_type %in% "city") %>%
  rename(mode = transportation_type) %>%
  select(country, region, mode, date, score) %>%
  duplicates(key = c(region, mode), index = date) %>%
  group_by(country, region) %>%
  tally()

```

Some cities have the same names. We have Dublin (Ireland) and Dublin (Georgia, as readers of the first page of _Finnegans Wake_ will recall). Similarly, Birmingham (UK and Alabama), Reading (UK and Pennsylvania), and Naples (Italy and Florida). This will mess up our attempt to create a `tsibble` uniquely keyed by city and mode of transportation.

There's some missing data we have to deal with, too, if we want to use some time series tools.

For now we'll just focus on some cities rather than every single one.

```{r}
focus_on <- c("Rio de Janeiro", "Lyon", "Bochum - Dortmund", "Dusseldorf",
              "Barcelona", "Detroit", "Toulouse", "Stuttgart",
              "Cologne", "Hamburg", "Cairo", "Lille", "New York")

apple_tsb <- apple_mobility %>%
  filter(region %in% focus_on) %>%
  rename(mode = transportation_type) %>%
  select(country, region, mode, date, score) %>%
  as_tsibble(key = c(region, mode), index = date) 

apple_tsb <- apple_tsb %>%
  model(arima = ARIMA(score ~ trend())) %>%
  interpolate(apple_tsb)

```

Now we can do some decompositions. In what follows we'll move back and forth between the base `apple_mobility` data (which includes missing values) and the `apple_tsb` object, which is a `tsibble` object that's had missing values interpolated, in preparation for the STL decomposition that we'll do.

```{r}

apple_tsb %>%
  model(STL(score)) %>%
  components() %>%
  ggplot(aes(x = date, y = trend, group = mode, color = mode)) +
  geom_line(size = 0.5) +
  scale_color_manual(values = my_colors("bly")) +
  facet_wrap(~ region, ncol = 8) +
  labs(x = "Date", y = "Trend",
       color = "Mode",
       title = "Driving, Selected Cities, Trend Component",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")
```

```{r}
## Raw
raw_ts <- apple_mobility %>%
  filter(geo_type == "city") %>%
  select(region:score) %>%
  rename(mode = transportation_type) %>%
  mutate(mode = tools::toTitleCase(mode),
         weekend = isWeekend(date),
         holiday = isHoliday(as.timeDate(date), listHolidays())) %>%
  filter(region %in% focus_on) %>%
  group_by(region) %>%
  mutate(max_day = ifelse(is_max(score), date, NA),
         max_day = as_date(max_day))


raw_ts %>%
  filter(!is.na(max_day))

p_raw <- ggplot(raw_ts, mapping = aes(x = date, y = score,
                       group = mode, color = mode)) +
  geom_vline(data = subset(raw_ts, holiday == TRUE),
             mapping = aes(xintercept = date),
             color = my_colors("bly")[5], size = 1.5, alpha = 0.1) +
  geom_hline(yintercept = 100, color = "gray40") +
  geom_line() +
  geom_text_repel(aes(label = format(max_day, format = "%a %b %d")),
                  size = rel(2), nudge_x = 1, show.legend = FALSE) +
  scale_color_manual(values = my_colors("bly")) +
  facet_wrap(~ region) +
  labs(x = "Date", y = "Relative Mobility",
       color = "Mode",
       title = "Relative trends in activity, selected cities. Baseline data with no correction for weekly seasonality",
       subtitle = "Data are indexed to 100 for each city's usage on January 13th 2020. Weekends shown as vertical bars. Date with highest relative activity score labeled.\nNote that in Apple's data 'Days' are defined as Midnight to Midnight PST.",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")

p_raw

ggsave(here("figures/raw.pdf"), p_raw, width = 15, height = 10)


p_raw_tall <- ggplot(raw_ts, mapping = aes(x = date, y = score,
                                      group = mode, color = mode)) +
  geom_vline(data = subset(raw_ts, holiday == TRUE),
             mapping = aes(xintercept = date),
             color = my_colors("bly")[5], size = 1.5, alpha = 0.1) +
  geom_hline(yintercept = 100, color = "gray40") +
  geom_line() +
  geom_text_repel(aes(label = format(max_day, format = "%a %b %d")),
                  size = rel(2), nudge_x = 1, show.legend = FALSE) +
  scale_color_manual(values = my_colors("bly")) +
  facet_wrap(~ region, ncol = 2) +
  labs(x = "Date", y = "Relative Mobility",
       color = "Mode",
       title = "Relative trends in activity, selected cities. No seasonal correction.",
       subtitle = "Data are indexed to 100 for each city's usage on January 13th 2020. Weekends shown as vertical bars.\nDate with highest relative activity index labeled.\nNote that in Apple's data 'Days' are defined as Midnight to Midnight PST.",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")

ggsave(here("figures/raw_tall.pdf"), p_raw_tall, width = 8, height = 12)
ggsave(here("figures/raw_tall.png"), p_raw_tall, width = 8, height = 12, dpi = 300)
```

```{r}
feats <- apple_mobility %>%
  filter(region %in% focus_on) %>%
  select(region:score) %>%
  rename(mode = transportation_type) %>%
  mutate(mode = tools::toTitleCase(mode)) %>%
  filter(region %in% focus_on) %>%
  as_tsibble(key = c(region, mode)) %>%
  features(score, feature_set(tags = "stl"))

resids <- apple_tsb %>%
  mutate(mode = tools::toTitleCase(mode)) %>%
  model(STL(score)) %>%
  components() %>%
  mutate(weekend = isWeekend(date),
         holiday = isHoliday(as.timeDate(date), listHolidays())) %>%
  as_tibble() %>%
  group_by(region) %>%
  mutate(max_day = ifelse(is_max(remainder), date, NA),
         max_day = as_date(max_day))

p <- ggplot(resids, aes(x = date, y = remainder, group = mode, color = mode)) +
  geom_vline(data = subset(resids, holiday == TRUE),
             mapping = aes(xintercept = date),
             color = my_colors("bly")[5], size = 1.5, alpha = 0.1) +
  geom_line(size = 0.5) +
  geom_text_repel(aes(label = format(max_day, format = "%a %b %d")),
                  size = rel(2), nudge_x = 1, show.legend = FALSE) +
  scale_color_manual(values = my_colors("bly")) +
  facet_wrap(~ region) +
  labs(x = "Date", y = "Remainder", color = "Mode",
       title = "Remainder component for activity data (after trend and weekly components removed)",
       subtitle = "Weekends shown as vertical bars. Date with highest remainder component labeled.\nNote that in Apple's data 'Days' are defined as Midnight to Midnight PST.",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")

ggsave("figures/remainders.pdf", p, width = 15, height = 10)

p_resid_tall <- ggplot(resids, aes(x = date, y = remainder, group = mode, color = mode)) +
  geom_vline(data = subset(resids, holiday == TRUE),
             mapping = aes(xintercept = date),
             color = my_colors("bly")[5], size = 1.5, alpha = 0.1) +
  geom_line(size = 0.5) +
  geom_text_repel(aes(label = format(max_day, format = "%a %b %d")),
                  size = rel(2), nudge_x = 1, show.legend = FALSE) +
  scale_color_manual(values = my_colors("bly")) +
  facet_wrap(~ region, ncol = 2) +
  labs(x = "Date", y = "Remainder", color = "Mode",
       title = "Remainder component for activity data (after trend and weekly components removed)",
       subtitle = "Weekends shown as vertical bars. Date with highest remainder component labeled.\nNote that in Apple's data 'Days' are defined as Midnight to Midnight PST.",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")


ggsave(here("figures/remainders_tall.pdf"), p_resid_tall, width = 8, height = 12)
ggsave(here("figures/remainders_tall.png"), p_resid_tall, width = 8, height = 12, dpi = 300)
```

```{r}
## NYC

raw_ny <- apple_mobility %>%
  filter(region == "New York City") %>%
  select(region:score) %>%
  rename(mode = transportation_type) %>%
  mutate(mode = tools::toTitleCase(mode),
         weekend = isWeekend(date),
         holiday = isHoliday(as.timeDate(date), listHolidays())) %>%
  mutate(max_day = ifelse(is_max(score), date, NA),
         max_day = as_date(max_day))

p_raw_ny <- ggplot(raw_ny, mapping = aes(x = date, y = score,
                                      group = mode, color = mode)) +
  geom_vline(data = subset(raw_ny, holiday == TRUE),
             mapping = aes(xintercept = date),
             color = my_colors("bly")[5], size = 2.9, alpha = 0.1) +
  geom_hline(yintercept = 100, color = "gray40") +
  geom_line() +
  geom_text_repel(aes(label = format(max_day, format = "%a %b %d")),
                  size = rel(2), nudge_x = 1, show.legend = FALSE) +
  scale_color_manual(values = my_colors("bly")) +
  labs(x = "Date", y = "Relative Mobility",
       color = "Mode",
       title = "New York City's relative trends in activity. Baseline data with no correction for weekly seasonality",
       subtitle = "Data are indexed to 100 for usage on January 13th 2020. Weekends shown as vertical bars. Date with highest relative activity score labeled.\nNote that in Apple's data 'Days' are defined as Midnight to Midnight PST.",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")

p_raw_ny

ggsave(here("figures/nyc_raw.pdf"), plot = p_raw_ny, height = 7, width = 10)
ggsave(here("figures/nyc_raw.png"), plot = p_raw_ny, height = 7, width = 10, dpi = 300)

resids_ny <- apple_tsb %>%
  filter(region == "New York City") %>%
  mutate(mode = tools::toTitleCase(mode)) %>%
  model(STL(score)) %>%
  components() %>%
  mutate(weekend = isWeekend(date),
         holiday = isHoliday(as.timeDate(date), listHolidays())) %>%
  as_tibble() %>%
  mutate(max_day = ifelse(is_max(remainder), date, NA),
         max_day = as_date(max_day))

p_resid_ny <- ggplot(resids_ny, aes(x = date, y = remainder, group = mode, color = mode)) +
  geom_vline(data = subset(resids, holiday == TRUE),
             mapping = aes(xintercept = date),
             color = my_colors("bly")[5], size = 2.9, alpha = 0.1) +
  geom_line(size = 0.5) +
  geom_text_repel(aes(label = format(max_day, format = "%a %b %d")),
                  size = rel(2), nudge_x = 1, show.legend = FALSE) +
  scale_color_manual(values = my_colors("bly")) +
  labs(x = "Date", y = "Remainder", color = "Mode",
       title = "New York City, Remainder component for activity data",
       subtitle = "Weekends shown as vertical bars. Date with highest remainder component labeled.\nNote that in Apple's data 'Days' are defined as Midnight to Midnight PST.",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")

ggsave(here("figures/nyc_remainders.pdf"), p_resid_ny, width = 10, height = 7)
ggsave(here("figures/nyc_remainders.png"), p_resid_ny, width = 10, height = 7, dpi = 300)
```

```{r}
trend_ny <- apple_tsb %>%
  filter(region == "New York City") %>%
  mutate(mode = tools::toTitleCase(mode)) %>%
  model(STL(score)) %>%
  components() %>%
  mutate(weekend = isWeekend(date),
         holiday = isHoliday(as.timeDate(date), listHolidays())) %>%
  as_tibble() %>%
  mutate(max_day = ifelse(is_max(remainder), date, NA),
         max_day = as_date(max_day))

p_trend_ny <- ggplot(trend_ny, aes(x = date, y = trend, group = mode, color = mode)) +
  geom_vline(data = subset(resids, holiday == TRUE),
             mapping = aes(xintercept = date),
             color = my_colors("bly")[5], size = 2.9, alpha = 0.1) +
  geom_line(size = 0.5) +
  scale_color_manual(values = my_colors("bly")) +
  labs(x = "Date", y = "Trend", color = "Mode",
       title = "New York City, trend component for activity data",
       subtitle = "Weekends shown as vertical bars.\nNote that in Apple's data 'Days' are defined as Midnight to Midnight PST.",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")

ggsave(here("figures/nyc_trend.pdf"), p_trend_ny, width = 10, height = 7)
ggsave(here("figures/nyc_trend.png"), p_trend_ny, width = 10, height = 7, dpi = 300)

```


```{r}
## NY vs SF vs Chicago vs LA
us_big_focus <- c("New York City", "Chicago", "Los Angeles", "Dallas", "Philadelphia", "Washington DC", "Atlanta", "San Francisco - Bay Area", "Boston")



## Raw
raw_ts <- apple_mobility %>%
  filter(region %in% us_big_focus) %>%
  select(region:score) %>%
  rename(mode = transportation_type) %>%
  mutate(mode = tools::toTitleCase(mode),
         weekend = isWeekend(date),
         holiday = isHoliday(as.timeDate(date), listHolidays())) %>%
  filter(region %in% focus_on) %>%
  group_by(region) %>%
  mutate(max_day = ifelse(is_max(score), date, NA),
         max_day = as_date(max_day))


p_raw_us <- ggplot(raw_ts, mapping = aes(x = date, y = score,
                                      group = mode, color = mode)) +
  geom_vline(data = subset(raw_ts, holiday == TRUE),
             mapping = aes(xintercept = date),
             color = my_colors("bly")[5], size = 1.5, alpha = 0.1) +
  geom_hline(yintercept = 100, color = "gray40") +
  geom_line() +
  geom_text_repel(aes(label = format(max_day, format = "%a %b %d")),
                  size = rel(2), nudge_x = 1, show.legend = FALSE) +
  scale_color_manual(values = my_colors("bly")) +
  facet_wrap(~ region) +
  labs(x = "Date", y = "Relative Mobility",
       color = "Mode",
       title = "Relative trends in activity, selected cities. Baseline data with no correction for weekly seasonality",
       subtitle = "Data are indexed to 100 for each city's usage on January 13th 2020. Weekends shown as vertical bars. Date with highest relative activity score labeled.\nNote that in Apple's data 'Days' are defined as Midnight to Midnight PST.",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")

p_raw_us

ggsave(here("figures/raw_us.pdf"), p_raw_us, width = 15, height = 10)
ggsave(here("figures/raw_us.png"), p_raw_us, width = 15, height = 10, dpi = 300)
```

Make a new `tsibble` object for US cities.

```{r}
us_big_tsb <- apple_mobility %>%
  filter(region %in% us_big_focus) %>%
  rename(mode = transportation_type) %>%
  select(country, region, mode, date, score) %>%
  as_tsibble(key = c(region, mode), index = date) 

## missing data again
us_big_tsb <- us_big_tsb %>%
  model(arima = ARIMA(score ~ trend())) %>%
  interpolate(us_big_tsb)

```


```{r}
resids <- us_big_tsb %>%
  mutate(mode = tools::toTitleCase(mode)) %>%
  model(STL(score)) %>%
  components() %>%
  mutate(weekend = isWeekend(date),
         holiday = isHoliday(as.timeDate(date), listHolidays())) %>%
  as_tibble() %>%
  group_by(region) %>%
  mutate(max_day = ifelse(is_max(remainder), date, NA),
         max_day = as_date(max_day))

p_us_resids <- ggplot(resids, aes(x = date, y = remainder, group = mode, color = mode)) +
  geom_vline(data = subset(resids, holiday == TRUE),
             mapping = aes(xintercept = date),
             color = my_colors("bly")[5], size = 1.5, alpha = 0.1) +
  geom_line(size = 0.5) +
  geom_text_repel(aes(label = format(max_day, format = "%a %b %d")),
                  size = rel(2), nudge_x = 1, show.legend = FALSE) +
  scale_color_manual(values = my_colors("bly")) +
  facet_wrap(~ region) +
  labs(x = "Date", y = "Remainder", color = "Mode",
       title = "Remainder component for activity data (after trend and weekly components removed)",
       subtitle = "Weekends shown as vertical bars. Date with highest remainder component labeled.\nNote that in Apple's data 'Days' are defined as Midnight to Midnight PST.",
       caption = "Data: Apple. Graph: @kjhealy") +
  theme(legend.position = "top")

ggsave(here("figures/remainders_us.pdf"), p_us_resids, width = 15, height = 10)
ggsave(here("figures/remainders_us.png"), p_us_resids, width = 15, height = 10, dpi = 300)


```

